---
title: "Data Handling: Group Project, Part 1"
subtitle: "Data Handling"
author:
  - name: Author1, Marco Molnar (student_id)
  - name: Author2, Lars Wyssen (student_id)
  - name: Author3, Alessio De Simeis (student_id)
  - name: Author4, Elia Locher (21-941-000)
date: "November 2025"
format:
  html:
    embed-resources: true
    self-contained: true
    self-contained-math: true
highlight-style: dracula
---

```{r setup, echo = FALSE, eval = TRUE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE)
```

# R Setup

```{r}
# Load necessary libraries
library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)
```

# Homework Part 1 (due November 6th, 2025)

The relationship between health expenditures and health outcomes has long been a central topic in health economics. Empirical studies show a positive but non-linear relationship between health spending and outcomes such as life expectancy, particularly in low- and middle-income countries.

In high-income settings, however, diminishing returns are evident. For example, the United States spends more per capita on health than any other nation, yet its health outcomes often lag behind those of countries with lower spending.

In this group project, we want to investigate the relationship between health expenditures and health outcomes.

## <br>

<br>

## Importing data on healthcare expenditures per capita

<br>

### Current health expenditure per capita and current health expenditure (% of GDP)

Here, write down your solution in text. Feel free to use standard markdown syntax for your text. For instance, *this*, or **this** or maybe you want to write code inline with `2+2`. For bullet points and lists, use

-   this
-   is
-   a bullet point (make sure you have one full empty line before the list starts)

Your code should be written in **code chunks**, with clear comments explaining each step. When you run the entire document, all objects created in the chunks are saved in your R environment.

If you want to run the code line by line instead, use the Run button or Ctrl + Enter (Cmd + Enter on Mac) to execute individual lines or selected code.

```{r}
# This is a vector. If I print it, it appears in the rendered document.
vec <- c(2,2)

# Not printed
```

```{r}
# This is a vector. If I print it, it appears in the rendered document.
vec <- c(2,2)
# Printed
print(vec)
```

```{r, eval = F}
# eval = FALSE: code chunk is shown, but not executed (no output is printed)
vec <- c(2,2)
vec
```

```{r, echo = F}
# echo = FALSE: code is executed, but the code itself is hidden (only output is printed)
vec <- c(2,2)
vec
```

```{r, echo = TRUE, results = 'hide'}
# results = 'hide': code is shown, executed, but output is hidden
vec <- c(2, 2)
vec
```

### What are the healthcare expenditures for Albania in 2005 in % of GDP?

<br>

### Plausibility checks

<br>

## Importing data on life expectancy and perceived health status for OECD countries

<br>

### Life expectancy at birth indicator

<br>

### Perceived Health Status

<br>

## First data exploration

### Content of your data

<br>

### Data dimensions and structure

<br>

## Save your data
