---
title: "Group Project Part "
subtitle: "3,230: Data Handling: Import, Cleaning and Visualisation - Autumn Semester 2025"
author:
  - name: Alessio De Simeis
  - name: Elia Valentin Locher
  - name: Marco Molnar
  - name: Lars Wyssen
date: "December 10, 2025"
format:
  html:
    embed-resources: true
    self-contained: true
    self-contained-math: true
    toc: true
    toc-location: left
    toc-depth: 6
    theme: cosmo
    code-fold: true
    code-tools: true 
highlight-style: dracula
editor: 
  markdown: 
    wrap: sentence
---

```{r setup, echo = FALSE, eval = TRUE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE)
```

# R Setup

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(skimr)
library(writexl)
library(ggrepel)

```

# Homework Part 2 (due December 10th, 2025)

The relationship between health expenditures and health outcomes has long been a central topic in health economics.
Empirical studies show a positive but non‐linear relationship between health spending and outcomes such as life expectancy, particularly in low‐ and middle‐income countries.

In high‐income settings, however, diminishing returns are evident.
For example, the United States spends more per capita on health than any other nation, yet its health outcomes often lag behind those of countries with lower spending (Anderson et al. (2003)).
In this group project, we investigate the relationship between health expenditures and health outcomes.

The objectives of part 2 are the following:

1.  Clean the data
2.  Merge the different dataframes
3.  Formulate a hypothesis to guide your research
4.  Explore the data
5.  Visualize the data to validate or invalidate your hypothesis.

## Import Datasets

First, we import the dataset which were given to have a good starting point.
We directly convert characters to factors.

```{r}
# Import data (change the path)
df_WorldBank_health <- read.csv("df_WorldBank_health.csv", stringsAsFactors = TRUE)

df_WorldBank_life_expectancy <- read.csv("df_WorldBank_life_expectancy.csv", stringsAsFactors = TRUE)

df_oecd_life_expectancy <- read.csv("df_oecd_life_expectancy.csv", stringsAsFactors = TRUE)

df_oecd_phs <- read.csv("df_oecd_phs.csv", stringsAsFactors = TRUE)
```

## 3. Data cleaning

::: {.callout-note title="Exercise"}
Explore the data to identify the relevant variables for your analysis, making sure their formats and values are correct.

-   Perform all the necessary cleaning steps, and clearly describe these steps in the code.
-   Comment on the cleaning choices you made regarding the data.
-   Keep only the variables of interest in your dataset.
-   Summarize your variables of interest.
:::

### 3.1 Health expenditures (World Bank)

#### Relevant Variables

First we will have a look at the health expenditures form the World Bank.

```{r}
str(df_WorldBank_health)
```

We can directly see that some variables are not relevant for our further analysis.
For each country and year we want the relevant number for health expenditure per capita in absolute USD and as percentage of its GDP.
Further, as stated in the exercise, we want to included the variable `income`.

**Therefore we have three variables of interest, as well as two variables which describes the country (iso3c and country, the latter one just for convenience) and year variable since we have time series**.
We first get rid of `X`, `iso2c`, `status`, `lastupdated`, `region`, `capital`, `longitude`, `latitude` and `lending`.
For the sake of simplicity, we rename two variables of interest:

```{r}
df_WorldBank_health <- df_WorldBank_health %>%
  select(-X, -iso2c, -status, -lastupdated, -region, -capital, -longitude, -latitude, -lending)

df_WorldBank_health <- df_WorldBank_health %>%
  rename(
    health_gdp = ind_WorldBank_health_gdp,
    health_usd = ind_WorldBank_health_usd
  )

head(df_WorldBank_health)
```

#### Cleaning Health Expenditures per Capita and as % of GDP

Now our dataset is much more compact and we have all variables we want.
Because time passed since we did the first part of the group project and we believe it is important to know our data well before diving deeper into the cleaning, we will have a quick look at the dataset using the recommended `skim()` function from the package `skimr`:

```{r}
skim(df_WorldBank_health)
df_WorldBank_health %>% 
  summarize(n_countries = n_distinct(country))
```

We see that:

-   we have 266 different countries in our dataset
-   For those countries, we have health expenditure data measured yearly ranging from 2000 up to 2022
-   health expenditure is measured in 1) per capita USD and 2) relative to the total GDP for the country in that year

Further, we can see that for the combination for some countries and year we have some missing values.
The number of missing values for our first variable of interest `health_gdp` and our second variable of interest `health_usd` are almost identical with 676 and 678 missing values, respectively.
The number of missing values for `income` is 276.
We investigate if countries with missing values in `health_usd` also have missing values in `health_gdp`

```{r}
matrix_2x2 <- df_WorldBank_health %>%
  mutate(
    health_usd_status = ifelse(is.na(health_usd), "Missing", "Not Missing"),
    health_gdp_status = ifelse(is.na(health_gdp), "Missing", "Not Missing")
  ) %>%
  count(health_usd_status, health_gdp_status) %>%
  tidyr::pivot_wider(
    names_from = health_gdp_status,
    values_from = n,
    values_fill = 0
  )

print(matrix_2x2)


```

This tibble shows what we would expect: `health_usd` and `health_gdp` are missing together, only in two cases `health_usd` is missing but `health_gdp` is not.
We don't want to rush and just handle these missing values without looking at it further, so let's see if we can find a pattern.

For this reason, we look for which countries we have missing values.
Is it like for some years we don't have values for a lot of countries or just for a small number of countries we don't have data at all?

```{r}
df_WorldBank_health %>%
  filter(is.na(health_usd)) %>%
  count(country, name = "n_missing_years") %>%
  arrange(desc(n_missing_years))

```

This is interesting!
For some countries like **American Samoa, Aruba, Bermuda and more** we have 23 missing values which means that we have completely no values at all (from 2000 up to 2022 there are 23 possible values including 2000).
Out of all missing values, how many are due to the fact that we don't have any data for some countries:

```{r}
# Count how many missing values per country
missing_by_country <- df_WorldBank_health %>%
  filter(is.na(health_usd)) %>%
  count(country, name = "n_missing")

# Identify countries with no data at all
countries_all_missing <- missing_by_country %>%
  filter(n_missing == 23)

n_countries_all_missing <- nrow(countries_all_missing)

# Out of all missing values for health_usd, how many can we explain with countries which have no values at all
missing_explained <- n_countries_all_missing * 23

# Total Missing values
total_missing <- df_WorldBank_health %>%
  summarise(n = sum(is.na(health_usd))) %>%
  pull(n)

# missing values explained in %
percent_explained <- (missing_explained / total_missing) * 100


cat(
  "\nNumber of countries with no data at all:", n_countries_all_missing,
  "\nTotal missing values", total_missing, sprintf("\nPercentage of explained missing values with countries no data at all: %.2f%%\n", percent_explained)
)

```

We can explain roughly 88% of all missing values with countries where we don't have any data at all.
In this case, we don't know exactly why we don't have any data for those countries.
Maybe there was a problem in the API Approach or the WorldBank just don't measure data for those countries.
We exclude those countries, because data imputation just doesn't make sense here:

```{r}
countries_all_missing <- df_WorldBank_health %>%
  group_by(country) %>%
  summarise(n_missing = sum(is.na(health_usd))) %>%
  filter(n_missing == 23) %>%
  pull(country)

df_WorldBank_health <- df_WorldBank_health %>%
  filter(!country %in% countries_all_missing)


```

Now we turn attention to **countries where we have missing values but not completely (meaning 23 NA's) missing.**

```{r}
missing_filtered <- missing_by_country %>%
  filter(n_missing < 23)

ggplot(missing_filtered, aes(x = n_missing)) +
  geom_histogram(binwidth = 1, color = "black", fill = "steelblue") +
  labs(
    title = "Distribtuion of Missing Values per Country (n_missing < 23)",
    x = "Number of Missing Values",
    y = "Number of Countries"
  ) +
  theme_minimal()

```

The bar chart above shows how many missing values each country has (excluding countries with all 23 values missing, which gives us 10 countries).
We can clearly observe two distinct groups:

**Countries with very few missing values (n_missing \< 5)**

```{r}
countries_low_missing <- missing_by_country %>%
  filter(n_missing < 5) %>%
  pull(country)

df_group_low_missing <- df_WorldBank_health %>%
  filter(country %in% countries_low_missing,
         is.na(health_usd)) %>%
  select(country, year, health_usd) %>%
  arrange(country, year)

df_group_low_missing
```

-   The inspection of countries with fewer than five missing observations shows that the **missing values occur almost exclusively at the beginning or end of the time period (Cuba which is in economic crises and Ukraine being in a war).** This pattern indicates that these countries did not report data in the early years or only started reporting recently, rather than having random gaps within their time series.

-   **Because the missing years are located at the boundaries of the dataset, imputation is not recommended.** Boundary gaps cannot be reliably interpolated since there is no preceding (for early years) or subsequent (for late years) information to support a meaningful estimation.

-   For this reason, the most reasonable approach is to **exclude only the missing years for these countries, while keeping the remaining valid data points.** This avoids introducing artificial trends and maintains the integrity of the observed time series.

```{r}
df_WorldBank_health <- df_WorldBank_health %>%
  filter(!(country %in% countries_low_missing & is.na(health_usd)))

```

**Countries with many missing values (n_missing ≥ 10)**

```{r}
countries_high_missing <- missing_by_country %>%
  filter(n_missing >= 10, n_missing < 23) %>%
  pull(country)

df_group_high_missing <- df_WorldBank_health %>%
  filter(country %in% countries_high_missing,
         is.na(health_usd)) %>%
  select(country, year, health_usd) %>%
  arrange(country, year)

df_group_high_missing

```

-   **For countries with more than ten missing observations, the data gaps occur in long consecutive blocks, typically at the beginning of the reporting period.**

-   Because these countries provide only a limited number of actual observations, and because the missing values are not randomly distributed but appear in extended stretches, **imputation is not appropriate.**

-   Filling these gaps would require fabricating entire time segments without any empirical basis.

-   Therefore, the most reasonable approach is to exclude these countries from analyses that rely on complete time series or to treat them separately as low-quality data.

```{r}
df_WorldBank_health <- df_WorldBank_health %>%
  filter(!(country %in% countries_high_missing))
```

#### Economic Status Variable Income Group

As seen before, we also have missing values for the economic status variable `income`.
Unfortunately, cleaning the above two variables regarding health expenditure did not fix this issue.
So let's have a look why we have NAs here:

```{r}
df_WorldBank_health %>% filter(is.na(income))
```

We see that we have some missing values, because the **unit of observation changed for those values**.
In general our dataset is structured in a long format where each row is one country in a given year.
But for the missing values, it seems that the data is aggregated.
Our unit of observation are now countries with **High income or low income** or aggregated countries such as **Latin America & Caribbean**.
As we don't want to change the unit of observation and the information of economic status is already contained in the variable `income`, we delete those observations.

```{r}
# Delete the aggreaged countries
df_WorldBank_health <- df_WorldBank_health %>%
  filter(!is.na(income)) %>%
  filter(income != "Aggregates")
```

And we check if the data cleaning worked and we have no missing values:

```{r}
df_WorldBank_health <- df_WorldBank_health %>%
  arrange(country, year)

df_WorldBank_health <- df_WorldBank_health |>
  mutate(
    income = factor(income, levels = c(
      "Low income",
      "Lower middle income",
      "Upper middle income",
      "High income"
    ))
  )

skim(df_WorldBank_health)
```

::: {.callout-note title="Summary Missing Values for Health Expenditure"}
Countries with all 23 values missing provide no usable information and are therefore excluded entirely from the analysis.
Countries with more than 10 missing observations show long consecutive gaps, usually at the beginning or end of the time period, which makes reliable imputation impossible; these countries are treated as low-quality data and removed from time-series analyses.

In contrast, countries with fewer than five missing values exhibit only small gaps, typically at the boundaries of the dataset.
Because boundary gaps lack sufficient surrounding information, imputation is not appropriate for these cases either.
Instead, the missing years for this group are simply omitted while retaining all valid observations.
This tiered approach ensures that the analysis remains consistent, transparent, and based on empirically supported data.
:::

### 3.2 Life Expectancy (World Bank)

Let's have a look at this dataset.
Our relevant variable of interest is life expecantcy at birth for the overall population.
We see that we have 23 missing values for `iso2c` and for `SP.DYN.LE00.IN` which is the variable of interest.
This number seems not to be random, since it is probably the case that we don't have values for one country.
Let's check that:

```{r}
str(df_WorldBank_life_expectancy)
skim(df_WorldBank_life_expectancy)
df_WorldBank_life_expectancy %>% filter(country == "Not classified")
```

And as we thought, one country is **not classified**.
Since we don't know which country this is, we can't merge these observations later with some other country.
Therefore we delete those observations.

```{r}
df_WorldBank_life_expectancy <- df_WorldBank_life_expectancy %>%
  filter(country != "Not classified")
```

But we still got some missing values for `iso2c` and we observe that we don't have this abbreviation for Namibia.
But this is not a problem, since we use `iso3c` as common identifier later to merge datasets.
So we can just delete this variable and don't lose any information at all.

```{r}
# Check for which country we don't have values for iso2c
df_WorldBank_life_expectancy %>% filter(is.na(iso2c))

# Delete variables which are NOT of interest
df_WorldBank_life_expectancy <- df_WorldBank_life_expectancy %>%
  mutate(year = as.integer(year)) %>%
  select(-X, -iso2c) %>%
  arrange(country, year)%>%
  rename(
    life_exp_WB = SP.DYN.LE00.IN
  )


# Data Cleaning Check
skim(df_WorldBank_life_expectancy)
```

### 3.3 Life Expectancy (OECD)

As always, we have a first look at the dataset using `str()` and `skim()`.
We directly notice that we have more variables in this dataset compared to the equivalent one from the World Bank.

```{r}
str(df_oecd_life_expectancy)
skim(df_oecd_life_expectancy)
```

We were asked to just keep life expectancy at birth for the overall population.
Although it would be really interesting to see the differences of life expectancy regarding gender, we do the following cleaning steps:

-   **Overall Population**: `SEX` = \_T
-   **Only keep relevant variables of interest**:
    -   `REF_AREA` as country identifier,
    -   `TIME_PERIOD` as year identifier,
    -   `ObsValue` as life expectancy at birth variable,
    -   
-   **Arrange the observations**: Showing countries and year in a chronological order
-   **Note**: Since we don't have any missing values for `ObsValue`, there is no further cleaning necessary.

```{r}
df_oecd_life_expectancy <- df_oecd_life_expectancy %>%
  filter(SEX == "_T") %>%
  select(REF_AREA, TIME_PERIOD, ObsValue, OBS_STATUS) %>%
  mutate(TIME_PERIOD = as.integer(TIME_PERIOD)) %>%
  arrange(REF_AREA, TIME_PERIOD)%>%
  rename(
    life_exp_from_oecd = ObsValue,
    status_oecd_life_exp = OBS_STATUS
  )

df_oecd_life_expectancy

```

### 3.4 Health Status (OECD)

Inspecting this dataset, we see that we have 18 variables.
We need to filter accordingly:

::: {.callout-note title="Perceived Health Status"}
We want to keep only those observations:

-   **Overall Population**: `SEX` = \_T
-   **aged 15 years old and over**: `AGE` = Y_GE15
-   **who report their health to be "good/very good (or excellent)**: `HEALTH_STATUS` = G

As we want to have one row for a country in a given year, we also do the following: - **keep overall socio-economic status**: `SOCIO_ECON_STATUS` = \_Z, We use the \_Z category because it represents the total population without any socio-economic disaggregation, providing the most complete and comparable dataset across countries.
This is essential for our analysis, as we later compare these indicators with health expenditure at an aggregated national level.

As we don't have any missing values for the variable of interest, we don't have to do further cleaning.
:::

```{r}
df_oecd_phs <- df_oecd_phs %>%
  filter(AGE == "Y_GE15", SEX == "_T", HEALTH_STATUS == "G", SOCIO_ECON_STATUS == "_Z") %>%
  select(REF_AREA, TIME_PERIOD, ObsValue, OBS_STATUS) %>%
  arrange(REF_AREA, TIME_PERIOD)%>%
  rename(
    phs_from_oecd = ObsValue,
    status_oecd_phs = OBS_STATUS
  )

skim(df_oecd_phs)
df_oecd_phs

```

We focus on 38 OECD countries

```{r}
oecd_iso3c <- c("AUS","AUT","BEL","CAN","CHL","COL","CRI","CZE","DNK","EST",
               "FIN","FRA","DEU","GRC","HUN","ISL","IRL","ISR","ITA","JPN",
               "KOR","LVA","LTU","LUX","MEX","NLD","NZL","NOR","POL","PRT",
               "SVK","SVN","ESP","SWE","CHE","TUR","GBR","USA")

df_oecd_phs <- subset(df_oecd_phs, REF_AREA %in% oecd_iso3c)
df_oecd_phs
```

## 4. Measuring healthcare quality

### 4.1 Proxies for quality

::: {.callout-note title="Definition of Proxy"}
-   A proxy variable is an observable variable that is used to represent an unobservable or difficult-to-measure concept based on a strong relationship with that concept.
    For example, IQ is often used as a proxy for individual ability when regressing grades on study time and controlling for ability.

-   In our case, life expectancy is a useful proxy for healthcare quality because it reflects long-term health outcomes, but it is also influenced by income, lifestyle, and environmental factors.
    Perceived health status reflects subjective well-being and access to care, but may be biased by individual perceptions and cultural differences.

Definition is based on: - [Sage](https://methods.sagepub.com/ency/edvol/the-sage-encyclopedia-of-social-science-research-methods/chpt/proxy-variable) - and our knowledge from Data Analytics II (Econometrics)
:::

### 4.2 Choosing between OECD and World Bank for life expectancy

We first explore, how many distinct countries we have in both datasets.

```{r}
df_WorldBank_life_expectancy <- df_WorldBank_life_expectancy |>
  rename(REF_AREA = iso3c,
         TIME_PERIOD = year)

# Explore unique REF_AREAs by dataset
df_country_counts <- bind_rows(
  df_oecd_life_expectancy %>% 
    distinct(REF_AREA) %>% 
    summarise(dataset = "OECD", n_countries = n()),
  
  df_WorldBank_life_expectancy %>%
    distinct(REF_AREA) %>% 
    summarise(dataset = "WorldBank", n_countries = n())
)

df_country_counts
```

```{r}
diff_OECD_WB <- df_oecd_life_expectancy |>
  left_join(
    df_WorldBank_life_expectancy,
    by = c("REF_AREA", "TIME_PERIOD")) |>
  select(TIME_PERIOD, REF_AREA, country, life_exp_from_oecd, life_exp_WB) |>
  mutate(diff_oecd_wb = life_exp_from_oecd - life_exp_WB,
         diff_status = if_else(abs(diff_oecd_wb) >= 0.1, "difference", ""))
```

Let's see the distribution of the differences.

```{r}
ggplot(diff_OECD_WB, aes(x = diff_oecd_wb)) +
  geom_histogram(bins = 100) +
  labs(
    title = "Distribution of OECD–World Bank Life Expectancy Differences",
    x = "Difference (OECD minus WB)",
    y = "Count"
  )

```

And get a better understanding, if the differences occure country specific.

```{r}
diff_OECD_WB %>%
  filter(diff_status == "difference") %>% # preserve only real differences
  group_by(REF_AREA) %>%
  summarise(
    n_differences = n() # number of rows with differences
  ) %>%
  arrange(desc(n_differences)) # sort for better interpretability

```

In 41 out of the 51 countries, we can observe differences in measurement.

We consulted the [Source and Methods Document](https://stats.oecd.org/wbos/fileview2.aspx?IDFile=3607ac98-7466-4ea2-9f96-5410382b1497) of the OECD as well as the Metadata JSON-file retrieved via this call:

We then compared both Sources alongside 3 categories:

**1. Data Scources**

`OECD`

-   Primarily uses Eurostat life tables for European countries and national statistical office data else
-   Many countries use abridged or full life tables, three-years averages, or country-specific adjustments (e.g. Australia, Belgium, Hungary, New Zealand)

These inputs reflect each country's own statistical practices

`World Bank`

-   Doesn't use national life tables directly
-   Instead, relies on the United Nations Population Division's World Population Prospects (WPP)

Thus, uses a harmonised global mortality dataset, not life table calculations from country-specific statistical bureaus

**2. Different methodological approaches**

`OECD`

-   calculations differ across countries
-   some series include breaks in time series (OBS_STATUS = B) or changes in methodology (OBS_STATUS = D)
-   some values are estimated or smoothed by the OECD itself
-   some countries use 3-year averages, not single-year values

These variations might create small but meaningful deviations in year-to-year estimates

`World Bank`

-   explains its life expectancy values are interpolated data from 5-year period data
-   annual values are mathematically interpolated

The resulting series might be smoother and less responsive to annual national revision

**3. Difference in smoothing and rounding**

`OECD`

-   values may be reported with fewer decimals and may include country-specific smoothing (e.g. Australia from 1995 3-year average)

`World Bank`

-   have more decimal precision and reflect interpolated estimates rather than directly observed mortality rates

::: {.callout-note title="Summary and Conclusion"}
`OECD estimates` are based on national life tables, potentially smoothed or averaged, and updated when national methods change.

`World Bank estimates` come from the UN-WPP, are globally harmonised (model-based), and include interpolated annual values (smoother).

Since our final dataset for the analyses only comprises OECD countries, we decided to work with the `OECD life expectancy`.
These values are based on offical national life tables provided by Eurostat or national statical offices and therefore reflect (in our opinion) the most accurate and up-to-date country specific mortality information.

However, for analyses that involve global comparison beyond OECD, the World Bank data might be more appropriate.
They provide a consistent methodological framework that enables cross-country comparability, even though they are model-based and interpolated.
:::

Note: Since we changed in the dataset df_WorldBank_life_expectancy the columnnames to REF_AREA and TIME_PERIOD, we need to rename them to previous names => reason: several people worked on this notebook :)

```{r}
df_WorldBank_life_expectancy <- df_WorldBank_life_expectancy |>
  rename(iso3c = REF_AREA,
         year = TIME_PERIOD)
```


## 5. Merge the data

In the following, **we merge our data sources to construct two final datasets for the subsequent analysis**.
We deliberately use two separate datasets rather than a single one **because the perceived health status indicator is only available for OECD countries, whereas the analyses in Exercises 6 and 7 compare healthcare expenditure and life expectancy across different income groups worldwide**.
Restricting these analyses to OECD countries only would substantially limit the scope of our study, as it would exclude low- and middle-income countries and therefore miss important cross-income differences.

### 5.1 World Bank Health Expenditure and Life expectancy

::: {.callout-tip title="Merge the data: World Bank Health Expenditure and Life expectancy"}
0.  Which Datasets do we merge?
    We merge **df_WorldBank_health** (healthcare expenditure) and **df_WorldBank_life_expectancy** (life expectancy from World Bank).

1.  Which variable(s) will you use to perform the merge?
    We use `country`, `iso3c` and `year` as common identifiers to merge both datasets.
    Note: Since these three variable are identical named in both datasets, it makes it easy to merge the two together.

2.  Are there any additional cleaning steps required before merging the datasets?
    No, since we do a left-join (see next point)

3.  What type of merge is appropriate?
    We conduct a left-join, because the countries in the dataset df_WorldBank_health is already cleaned for countries and doesn't contain any aggregate levels.

4.  Are there any steps necessary in cleaning data after the merge?
    No, because we don't have any missing values - our cleaning approach in step 3 was good enough ;-)
:::

```{r}
df_WorldBank <- df_WorldBank_health %>%
  left_join(
    df_WorldBank_life_expectancy,
    by = c("country","iso3c","year"))

skim(df_WorldBank)
```

### 5.2 Perceived Health Status for OECD countries

::: {.callout-tip title="Merge the data: Perceived Health Status for OECD countries"}
0.  Which Datasets do we merge?

-   First, we merge **df_WorldBank_health** (healthcare expenditure) with **df_WorldBank_life_expectancy** (life expectancy from World Bank) to **df_WorldBank**. As this is the same as before, we can use the merged dataset from 5.1.
-   Second, we merge **df_WorldBank** with **df_oecd_phys** to **df_oecd_merged**.

1.  Which variable(s) will you use to perform the merge?
    We now have the situation where we have two common identifiers, but these two variables are named differently in the datasts.
    So we use `REF_AREA`(df_oecd_phs) and `iso3c` (df_WorldBank) as well as `TIME_PERIOD` (df_oecd_phs) and `year` to merge both datasets.

2.  Are there any additional cleaning steps required before merging the datasets?
    No, but we will have to do some cleaning after merging (see below).

3.  What type of merge is appropriate?
    We conduct a full_join, because we first want to have all observations from both datasets available and then do some cleaning afterwards if necessary (spoiler, it is necessary).

4.  Are there any steps necessary in cleaning data after the merge?

-   Yes. We see that for a majority of observations, we don't have data for both healthcare quality proxies.
-   For 3,667 observations, we only have the numbers for life expectancy, but not for perceived health status, as we use data for life expectancy from the World Bank (see explanation exercise 4.2) and health status for only OECD countries.
-   As described above, with our two final dataset approach, we will delete in this merged dataset all observations where we don't have any data for perceived health status and end up with a dataset for only OECD countries.
:::

```{r}
df_oecd_merged <- df_oecd_phs %>%
  full_join(
    df_WorldBank,
    by = c("REF_AREA" = "iso3c",
       "TIME_PERIOD" = "year"
    ))

nrow(df_oecd_merged %>% filter(!is.na(life_exp_WB) & !is.na(phs_from_oecd)))
nrow(df_oecd_merged %>% filter(!is.na(life_exp_WB) & is.na(phs_from_oecd)))

df_oecd_merged_final <- df_oecd_merged %>%
  filter(!is.na(life_exp_WB) & !is.na(phs_from_oecd))
```

## 6. Explore and visualize the data

### 6.1 Trends in expenditures

#### Summary statistics of key variables

-   **What we did**: We computed summary statistics for `health_usd` and `health_gdp` to obtain a first overview of their distributions.

-   **What we see**: Per-capita health expenditure in USD shows very large dispersion, ranging from about 4 USD to over 12,000 USD, indicating strong cross-country inequality.
    In contrast, health expenditure as a share of GDP is much more concentrated, with a mean of about 6.2% and a maximum of around 24%.

```{r}
# First quick look at distributions of both relevant variables
summary(df_WorldBank$health_usd)
summary(df_WorldBank$health_gdp)
```

#### Yearly aggregation of global health expenditure

-   **What we did**: We grouped the data by year and computed the yearly mean of per-capita health expenditure in USD, the mean health expenditure as a share of GDP, and the number of available countries per year.

-   **What we see**: From 2000 to 2022, average health spending increased from about 443 USD to over 1,360 USD per capita, while the average health expenditure share of GDP remained relatively stable around 5.5%–7%.
    The number of observed countries per year is stable at around 225, indicating consistent global coverage over time.

```{r}
health_trends <- df_WorldBank |>
  group_by(year) |>
  summarise(
    avg_health_usd = mean(health_usd, na.rm = TRUE),
    avg_health_gdp = mean(health_gdp, na.rm = TRUE),
    n_countries = n()
  )

summary(health_trends)

```

#### Trend in global healthcare spending per capita (USD)

-   **What we did**: We visualized the yearly average of per-capita healthcare expenditure in USD using a line plot to analyze its evolution over time.

-   **What we see**: Global average health spending per capita increases strongly from around 450 USD in 2000 to over 1,350 USD by 2021, with a noticeable dip around 2014–2015 and a sharp rise after 2019.
    This indicates a strong long-term upward trend in nominal healthcare expenditure.

```{r}
ggplot(health_trends, aes(x = year, y = avg_health_usd)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Global Average Healthcare Expenditure per Capita (USD)",
    x = "Year",
    y = "USD per Capita"
  ) +
  theme_minimal()

```

#### Trend in global healthcare spending as a share of GDP

-   **What we did**: We visualized the yearly average of healthcare expenditure as a percentage of GDP using a line plot to analyze its evolution over time.

-   **What we see**: On average, health expenditure as a share of GDP increases only moderately from about 5.5% in 2000 to around 7–7.2% in 2020–2021, followed by a slight decline in 2022.
    Compared to the steep rise in spending in USD, this indicates a much flatter relative trend.

```{r}
ggplot(health_trends, aes(x = year, y = avg_health_gdp)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Global Average Healthcare Expenditure as % of GDP",
    x = "Year",
    y = "Percent of GDP"
  ) +
  theme_minimal()

```

#### Caution when interpreting the trend

When interpreting the strong increase in healthcare expenditure per capita in USD, it is important to note ...

-   ... that these values are expressed in nominal terms and therefore partly reflect inflation and general economic growth, not only real increases in healthcare provision.
-   In addition, global averages conceal substantial heterogeneity across countries, meaning that the upward trend is mainly driven by high-income countries while many low-income countries experience much slower growth (see 6.2).
-   Finally, the relatively flat evolution of healthcare spending as a share of GDP shows that rising absolute spending does not necessarily imply a proportional increase in the economic importance of healthcare.

#### Summary

::: {.callout-note title="Summary 6.1 Trends in expenditures"}
While healthcare expenditure per capita in USD shows a strong and continuous upward trend, the corresponding share of GDP increases only moderately over the same period.
This indicates that the sharp rise in absolute spending is driven not only by a rising importance of healthcare, but also by general economic growth and price inflation.
When interpreting these trends, it is important to keep in mind that the USD values are nominal and that global averages may conceal large inequalities between countries.
:::

### 6.2 Expenditures per income group

#### Summary statistics of key variable

-   **What we did**: We recoded the variable `income` as an ordered factor following the correct economic ranking from low to high income.

-   **What we see**: Most observations belong to high-income countries (1,426), followed by upper-middle-income (1,190), lower-middle-income (1,147), and low-income countries (527), indicating a stronger data coverage for richer economies.

```{r}
summary(df_WorldBank$income)
```

#### Which income group shwos the steepest increase in per-capita health speding?

-   **What we did**: We computed the yearly average of per-capita healthcare expenditure in USD separately for each income group.

-   **What we see**: After correctly ordering the income groups and analyzing the trends over time, we find that per-capita healthcare expenditure increases in all income groups, but at very different levels and speeds.
    High-income countries show by far the steepest increase in per-capita health spending over the last decades, while low- and lower-middle-income countries remain at much lower spending levels throughout the period.
    This highlights the persistent and widening inequality in healthcare expenditure across income groups.

```{r}
health_income_trends <- df_WorldBank |>
  group_by(year, income) |>
  summarise(
    avg_health_usd = mean(health_usd, na.rm = TRUE),
    n_countries = n(),
    .groups = "drop"
  )

ggplot(health_income_trends, aes(x = year, y = avg_health_usd, color = income)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Healthcare Expenditure per Capita by Income Group",
    x = "Year",
    y = "USD per Capita",
    color = "Income group"
  ) +
  theme_minimal()
```

### 6.3 Life Expectancy

#### Quick check of life expectancy

-   **What we did**: We inspected summary statistics of `life expectancy` to obtain a first overview of the variable.

-   **What we see**: Life expectancy ranges from about 15 to 86 years, with a median of roughly 71.5 years and a mean of about 70 years, indicating substantial variation in mortality outcomes across countries and time.

```{r}
summary(df_WorldBank$life_exp_WB)
```

#### Aggregate life expectancy by year and income group

-   **What we did**: We visualized the yearly average life expectancy for each income group using a line plot to compare trends across low-, lower-middle-, upper-middle-, and high-income countries.

-   **What we see**: Life expectancy increases over time for all income groups, but at different levels and speeds: high-income countries consistently exhibit the highest life expectancy, while low-income countries show the lowest levels but also substantial improvements.
    The gap between low- and high-income groups narrows slightly over time, indicating partial convergence, although large differences persist.

```{r}
life_income_trends <- df_WorldBank |>
  group_by(year, income) |>
  summarise(
    avg_life_exp = mean(life_exp_WB, na.rm = TRUE),
    n_countries  = n(),
    .groups = "drop"
  )

ggplot(life_income_trends, aes(x = year, y = avg_life_exp, color = income)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Life Expectancy by Income Group",
    x = "Year",
    y = "Life Expectancy (years)",
    color = "Income group"
  ) +
  theme_minimal()
```

### 6.4 Perceived health status

#### Quick summary of perceived health

-   **What we did**: We inspected summary statistics of the perceived health status variable to obtain a first overview of its distribution.
-   **What we see**:

```{r}
summary(df_oecd_phs$phs_from_oecd)
```

#### Average perceived health status over time (OECD)

-   **What we did**: We computed and visualized the yearly average perceived health status across OECD countries using a line plot to analyze its evolution over time.

-   **What we see**: Average perceived health fluctuates around a relatively stable level between roughly 66 and 73 over the period 2000–2022, with no strong long-term upward or downward trend, but noticeable short-term variations.

```{r}
health_status_trend <- df_oecd_phs |>
  group_by(TIME_PERIOD) |>
  summarise(
    avg_health_status = mean(phs_from_oecd, na.rm = TRUE),
    n_countries = n(),
    .groups = "drop"
  )

ggplot(health_status_trend, aes(x = TIME_PERIOD, y = avg_health_status)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Average Perceived Health Status (OECD)",
    x = "Year",
    y = "Average perceived health status"
  ) +
  theme_minimal()

```

#### Cross-country differences (Boxplot)

-   **What we did**: We visualized the yearly distribution of perceived health status across OECD countries using boxplots to assess cross-country dispersion and the presence of outliers over time.

-   **What we see**: Perceived health varies substantially across OECD countries in every year, with a wide interquartile range and several low outliers, indicating persistent cross-country differences despite relatively stable OECD-wide averages over time.

```{r}
ggplot(df_oecd_phs, aes(x = factor(TIME_PERIOD), y = phs_from_oecd)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Perceived Health Status Across OECD Countries",
    x = "Year",
    y = "Perceived health status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))

```

::: callout-important
Since perceived health status is only available for OECD countries, the results are not representative of global health conditions but reflect patterns within high-income economies only.
:::

::: {.callout-tip title="Summary 6.4 Perceived Health Status"}
Perceived health status is analyzed for OECD countries only and therefore reflects self-reported health in high-income economies.
On average, perceived health remains relatively stable over time, with only moderate changes across years.
At the same time, there are sizable differences between OECD countries, indicating that subjective health perceptions vary strongly despite similar income levels
:::

## 7. Three Hypotheses

### 7.1 Formulate your hypotheses

we’re interested in the correlation between healthcare expenditure and healthcare quality at the country level.
More specifically, we want to test three hypotheses.
To guide you in the formulation of these three hypotheses, answer the following questions without using the data, be as concise as possible.

1.  Ex ante, would you expect a correlation between healthcare expenditure and healthcare quality? If so, would it be a positive or negative correlation?

**We expect a positive but moderate correlation between healthcare expenditure and healthcare quality in OECD countries, with its strength depending on the proxy for healthcare quality used.**

> Reason: Higher spending generally enables greater access to medical resources, technologies, and services, which should improve health outcomes.
> However, due to diminishing marginal returns to healthcare spending in high-income settings (most OECD countries) and the fact that higher expenditures may partly reflect higher prices rather than greater service provision, this positive relationship is expected to be moderate rather than strong in our data sample.
> Moreover, the strength of the association depends on the chosen proxy for healthcare quality, as indicators like life expectancy and perceived health status capture distinct dimensions of health that may respond differently to increased spending (see Hypthesis 2).

2.  Do you think health status and life expectancy are correlated? If so, would it be a positive or negative correlation?

**We expect a positive but weak correlation between perceived health status and life expectancy.**

> Reason: Although both perceived health status and life expectancy reflect aspects of population health, they capture fundamentally different dimensions: perceived health status is subjective and culturally influenced, while life expectancy is an objective, long-term outcome.
> Because perceived health status is recorded by a survey, people may report feeling healthy despite underlying risks, or conversely underrate their health due to cultural norms or expectations.

3.  Do you expect the correlation between expenditures and healthcare quality to be the same for high and low income countries?

**We expect the correlation between healthcare expenditure and healthcare quality to be stronger in low-income countries and weaker in high-income countries.**

> Reason: In low-income countries, additional health spending often translates directly into improved access to essential services, leading to substantial gains in health outcomes.
> In contrast, high-income countries operate in a range of diminishing marginal returns, where higher expenditures frequently reflect higher prices or inefficiencies rather than better care.

### 7.2 Validate (or invalidate) your three hypotheses

##### 7.2.1 Correlation between healthcare expenditure and healthcare quality

**What we did**:

- We created four scatterplots for 2019 to examine how healthcare expenditure relates to healthcare quality. We used two proxies for quality: `perceived health status` and `life expectancy`. And plotted each against both `health expenditure per capita` and `health expenditure as % of GDP`, adding linear regression lines to visualize the relationship.

**What we see**:

- All four plots show a *positive but moderate* association: higher spending is generally linked to better perceived health and higher life expectancy. The relationship is *clearer* for per-capita expenditure than for expenditure as a share of GDP, and perceived health displays a more linear trend than life expectancy, which shows *diminishing returns* at high spending levels. This supports our hypothesis of a positive but moderate relationship, with stronger associations for absolute spending levels than for relative GDP shares.

::: {.callout-note title="Correlation between healthcare expenditure and healthcare quality"}
The results confirm our hypothesis: the correlation between expenditure and healthcare quality is positive yet moderate, and its strength varies depending on the chosen health indicator.
:::

```{r}
#filter dataframes for 2019
merged19 <- df_oecd_merged_final |> dplyr::filter(TIME_PERIOD == 2019) # only OECD countries because of variable health status
wb19 <- df_WorldBank |> dplyr::filter(year == 2019) # all countries

#stile
theme_clean <- theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey90"),
    panel.grid.major.y = element_line(color = "grey90"),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 15),
    axis.title = element_text(face = "bold"),
    axis.text  = element_text(color = "grey20"),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  )

#scatterplot 1: perceived health status - helath expenditure (per capita in USD) (
ggplot(merged19,
       aes(x = health_usd, y = phs_from_oecd)) +
  geom_point(size = 2.2, alpha = 0.8, color = "grey20") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 1.1) +
  geom_text_repel(aes(label = REF_AREA),
                  size = 3,
                  max.overlaps = 25,) +
  labs(
    title = "Perceived Health Status and Health Expenditure per Capita 2019",
    x = "Health Expenditure per Capita (USD)",
    y = "Perceived Health"
  ) +
  theme_clean

#scatterplot 2: perceived health status - health expenditure (% of GDP)
ggplot(merged19,
       aes(x = health_gdp, y = phs_from_oecd)) +
  geom_point(size = 2.2, alpha = 0.8, color = "grey20") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 1.1) +
  geom_text_repel(aes(label = REF_AREA),
                  size = 3,
                  max.overlaps = 25) +
  labs(
    title = "Perceived Health Status and Health Expenditure (% of GDP) 2019",
    x = "Health Expenditure (% of GDP)",
    y = "Perceived Health"
  ) +
  theme_clean

#only label "important" countries
wb19_highlight <- wb19 |>
  filter(
    health_usd >= quantile(health_usd, 0.9, na.rm = TRUE) |
      health_usd <= quantile(health_usd, 0.1, na.rm = TRUE)
  )

#scatterplot 3: Life expectancy - health expenditure (per capita in USD)
ggplot(wb19,
       aes(x = health_usd, y = life_exp_WB)) +
  geom_point(size = 1.5, alpha = 0.6, color = "grey50") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 1.1) +
  geom_point(data = wb19_highlight,
             aes(x = health_usd, y = life_exp_WB),
             size = 2.2, color = "black") +
  geom_text_repel(
    data = wb19_highlight,
    aes(label = iso3c),
    size = 3,
    max.overlaps = 50
  ) +
  labs(
    title = "Life Expectancy and Health Expenditure per Capita 2019",
    x = "Health Expenditure per Capita (USD)",
    y = "Life Expectancy at Birth (years)"
  ) +
  theme_clean

#only label "important" countries
wb19_highlight_gdp <- wb19 |>
  filter(
    health_gdp >= quantile(health_gdp, 0.9, na.rm = TRUE) |
      health_gdp <= quantile(health_gdp, 0.1, na.rm = TRUE)
  )

#scatterplot 4: Life expectancy - health expenditure (% of GDP) (WorldBank_cleaned)
ggplot(wb19,
       aes(x = health_gdp, y = life_exp_WB)) +
  geom_point(size = 1.5, alpha = 0.6, color = "grey50") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 1.1) +
  geom_point(data = wb19_highlight_gdp,
             aes(x = health_gdp, y = life_exp_WB),
             size = 2.2, color = "black") +
  geom_text_repel(
    data = wb19_highlight_gdp,
    aes(label = iso3c),
    size = 3,
    max.overlaps = 50,
  ) +
  labs(
    title = "Life Expectancy and Health Expenditure (% of GDP) 2019",
    x = "Health Expenditure (% of GDP)",
    y = "Life Expectancy at Birth (years)"
  ) +
  theme_clean
```

##### 7.2.2 Correlation between healthcare expenditure and healthcare quality: a more precise answer

**What we did**:

- We quantified the relationship between healthcare expenditure and healthcare quality using the *Pearson correlation coefficient*, calculated with the R function cor(). Pearson’s r measures the *strength and direction of a linear association* between two continuous variables (range: –1 to +1).

- We computed correlations between both expenditure measures (`health_usd`, `health_gdp`) and both quality indicators (`perceived health status`, `life expectancy`) for the year 2019.

**What we see**:

- The results show consistently *positive but varying correlations*, depending on both the expenditure measure and the quality proxy:
  - health_usd vs. perceived health: 0.468 → moderate
  - health_gdp vs. perceived health: 0.308 → moderate
  - health_usd vs. life expectancy: 0.604 → moderately strong
  - health_gdp vs. life expectancy: 0.196→ weak
  
- Two clear patterns emerge:
1. `Per-capita expenditure (health_usd)` shows *stronger correlations* than expenditure as `% of GDP`.
2. `Perceived health status` displays a *more linear and stronger association* than `life expectancy`, which is more influenced by structural and long-term factors.

```{r}
# OECD perceived health
cor_phs_usd  <- cor(merged19$health_usd, merged19$phs_from_oecd, use = "complete.obs")
cor_phs_gdp  <- cor(merged19$health_gdp, merged19$phs_from_oecd, use = "complete.obs")

# World Bank life expectancy
cor_life_usd <- cor(wb19$health_usd, wb19$life_exp_WB, use = "complete.obs")
cor_life_gdp <- cor(wb19$health_gdp, wb19$life_exp_WB, use = "complete.obs")

# Clean formatted output
cat("Correlation results (2019):\n")
cat("--------------------------------------------------\n")
cat("Healthcare exp. USD per capita vs. Perceived health (OECD):", round(cor_phs_usd, 3), "\n")
cat("Healthcare exp. % of GDP vs. Perceived health (OECD):", round(cor_phs_gdp, 3), "\n\n")

cat("Healthcare exp. USD per capita vs. Life expectancy (World Bank):", round(cor_life_usd, 3), "\n")
cat("Healthcare exp. % of GDP vs. Life expectancy (World Bank):", round(cor_life_gdp, 3), "\n")

```

##### 7.2.3 Correlation between health status and life expectancy

**What we did**:

- We created a scatter plot with perceived health status on the x-axis and life expenditure from the World Bank on the y-axis.

**What we see**: 

- Our analysis of the correlation between life expectancy and perceived health status across OECD countries in 2019 shows a moderate positive correlation of 0.37. This indicates that countries in which a larger share of the population reports being in good or very good health also tend to have higher life expectancy.

- **This finding partially contradicts our initial hypothesis**, in which we expected only a weak relationship due to conceptual differences between the two measures. Although perceived health status is subjective and culturally influenced, while life expectancy is an objective long-term demographic outcome, the empirical results demonstrate that the two indicators are more closely related than anticipated.

- Overall, the analysis suggests that self-reported health is a meaningful predictor of broader population health outcomes across OECD countries.


```{r}
# Select only relevant columns
df_2019_phs_le <- merged19 |>
  select (REF_AREA, TIME_PERIOD, phs_from_oecd, life_exp_WB)

# Create scatter plot
ggplot(df_2019_phs_le, aes(x = phs_from_oecd, y = life_exp_WB)) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue", size = 1, alpha = 0.5) +
  geom_point(size = 3, alpha = 1, color = "darkblue") +
  geom_text(aes(label = REF_AREA), vjust = 2.5, size = 2 ) +
  labs(
    title = "Correlation between health status and life expectancy (OECD, 2019)",
    x = "Perceived Health status",
    y = "Life Expectancy"
  ) +
  theme_minimal()

rel_phs_le <- cor(df_2019_phs_le$phs_from_oecd, df_2019_phs_le$life_exp_WB)
cat(
  "\nCorrleation between health status and life expectancy for OECD countries:", rel_phs_le
)
```

##### 7.2.4 Correlation between expenditures and healthcare quality for high and low income countries?

**What we see**:

- **USD/life expectancy**: Within each income group, there is a positive correlation between health expenditure in USD and life expectancy. This relationship is strongest in low- and lower-middle-income countries, and weaker in high-income countries. This suggests that an increase in health spending can lead to large gains in life expectancy in low-income countries, whereas even large differences in spending lead to small differences in life expectancy in high-income countries.

- **GDP/life expectancy**: The relationship between health spending as % of GDP and life expectancy is weak and sometimes even slightly negative especially in middle income groups. This means that even if countries spend a larger share of GDP on health doesn't necessarily result in a higher life expectancy.


```{r}
ggplot(wb19 |> filter(income %in% c("High income", "Upper middle income", "Lower middle income", "Low income")),
       aes(x = health_usd, y = life_exp_WB, color = income)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  facet_wrap(~ income, scales = "free") +
  geom_text(aes(label = iso3c), vjust = 2.5, size = 2 ) +
  labs(
    title = "Health Expenditure (USD) vs Life Expectancy (2019)",
    x = "Health Expenditure (USD)",
    y = "Life Expectancy",
    color = "Income Group"
  ) +
  theme_minimal()

ggplot(wb19 |> filter(income %in% c("High income", "Upper middle income", "Lower middle income", "Low income")),
       aes(x = health_gdp, y = life_exp_WB, color = income)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  facet_wrap(~ income, scales = "free") +
  geom_text(aes(label = iso3c), vjust = 2.5, size = 2 ) +
  labs(
    title = "Health Expenditure (GDP) vs Life Expectancy (2019)",
    x = "Health Expenditure (GDP)",
    y = "Life Expectancy",
    color = "Income Group"
  ) +
  theme_minimal()

```

##### 7.2.5 Final graph

**What we did**:

- Here, we grouped the data frame by income and year. We then took the mean for each country and year for health expenditure and life expectancy. By grouping them by income group, we can later see how life expectancy reacts to health expenditure in USD per income group.

**What we see**:

- There is a clear positive relationship between health expenditure per capita (USD) and life expectancy over the years. At low and middle expenditure levels, life expectancy increases relatively quickly, but at high expenditure levels, the curve flattens, which describes the diminishing marginal returns as expenditure increases (we just have to be careful not to interpret here causality).

- Income groups are clearly separated: high-income countries cluster at high expenditure and high life expectancy, while upper-middle-income countries are located at lower expenditure and lower life expectancy.

- This pattern is a textbook example of diminishing marginal returns. In lower-income settings, additional healthcare spending is far more beneficial than in settings where most basic needs are already met. Therefore, we can conclude that more spending still helps, but each additional dollar buys less extra life expectancy.

```{r}
# Aggregate by income group and year (as required)
df_group <- df_WorldBank |>
  group_by(income, year) |>
  summarise(
    mean_health_usd  = mean(health_usd, na.rm = TRUE),
    mean_life_exp_WB = mean(life_exp_WB, na.rm = TRUE),
    .groups = "drop"
  )

# Keep only the four World Bank income groups
df_group_filtered <- df_group |>
  filter(income %in% c("Low income",
                       "Lower middle income",
                       "Upper middle income",
                       "High income"))

# Create final graph: Life expectancy vs expenditure (USD) by income group over time
ggplot(df_group_filtered,
       aes(x = mean_health_usd,
           y = mean_life_exp_WB,
           color = income,
           group = income)) +
  geom_path(size = 1) +     # evolution over time
  geom_point(size = 2) +
  labs(
    title = "Evolution of Life Expectancy and Health Expenditure by Income Group",
    x = "Average health expenditure per capita (USD)",
    y = "Average life expectancy (years)",
    color = "Income group"
  ) +
  theme_minimal()

```



## 8. Wrap up

### 8.1 Your data story

Our project tells a simple story: **how far does one extra dollar of health spending really go?**

Using World Bank and OECD data for 2000–2022, we cleaned and merged global series on health expenditure, life expectancy and perceived health. We show that spending per capita has more than tripled worldwide, yet gains in life expectancy are far from proportional. Low- and lower-middle-income countries still spend little, but every additional dollar is associated with large improvements in longevity.

In contrast, high-income countries already sit on a “plateau”: they spend vastly more, but extra spending buys only small health gains and often reflects higher prices rather than more care. Perceived health in OECD countries rises with spending too, but only moderately, reminding us that subjective well-being and objective survival are related yet distinct.

**Overall, our data story is about inequality and diminishing returns**: health systems in poor countries need more resources, while rich countries need to spend smarter rather than simply more.


#### 8.1.1 Further steps

-   **Additional methodology**: Our analysis is correlational; future work could study causal effects using instrumental variables or difference-in-differences designs.

-   **New research question**: Further, we could analyze why health spending improves health much more in low-income than in high-income countries (access, prevention, technology).

-   **Further data quality checks**: We could deepen our analysis by systematically examining OECD data breaks (B: Time_Series_Break, E: Estimated_Values, D: Definition_differs)